---
title: Calculating time intervals in working days using a calendar table
author: James Gill
date: '2020-11-22'
slug: calculating-time-intervals-in-working-days-using-a-calendar-table
categories:
  - SQL
tags:
  - Early career
  - SQL
  - T-SQL
summary: 'Yet another good reason to maintain a calendar table.'
output:
  blogdown::html_page:
    df_print: paged
---

*All data in this post is fake data, generated artificially.*

```{r setup, include=FALSE}
library(odbc)

conmssql <- dbConnect(odbc(),
                 Driver   = "SQL Server",
                 Server   = "DESKTOP-5R64MMS\\SQLEXPRESS",
                 Database = "SynthNHS",
                 Trusted_Connection = "True")
                 
knitr::opts_chunk$set(connection = "conmssql")                 
```

```{sql prepareexample, include=FALSE, code=readr::read_file("2020-11-22-prepare-example.sql")}
#prepare example table
```

I sometimes want to look at waiting times or turnaround times measured only in working days. This might be when thinking about a service which is only open on working days. Sometimes I've also seen service-level agreements specify performance metrics measured in working days.

A calendar table - a table in which each record represents a single calendar day - gives you a neat way of doing that.

The code looks like this:

```{sql wdturnaround, output.var = "wdturnaroundtbl", results = 'hide', code=readr::read_file("2020-11-22-turnaround-example.sql")}

```
``` {r wdturnaroundprint, echo = FALSE, rows.print = 20}
wdturnaroundtbl
```

The key is the correlated sub-query, which returns a row for each day in the calendar table (`lkp_calendar`) that's labelled as a working day between the request date and the completion date. The dates in the calendar table don't have a time, so SQL Server interprets them as midnight on the date in question. Therefore the code counts the midnights between the two dates, as long as the following day is a working day. 

In particular, requests made on a Friday and completed on Saturday or Sunday have a turnaround of 0 days, while requests made on a Sunday and completed on a Monday have a turnaround of 1 day. The behaviour for these edge cases can be adjusted as desired in the sub-query using `CASTS` and choosing between strict and non-strict inequalities.

There are lots of alternative approaches out there for counting working days. I like this approach because it's a clean way to account for bank holidays. I also find it easier to reason about our calendar table, which is very familiar to me, than I do to reason about some of the more obscure date functions in T-SQL.